Conditional resampling improves calibration \\ in single cell CRISPR screen analysis

### 2020-11-28 sceptre_paper/manuscript/manuscript_v3.tex
update manuscript

\usepackage{graphicx}	
\usepackage{amssymb, amsmath, bbm}
\usepackage{geometry}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\newcommand\independent{\perp \!\!\! \perp}
\usepackage{chngcntr}

@@

\begin{abstract}
Single cell CRISPR screens are an emerging biotechnology promising unprecedented insights into gene regulation. However, the analysis of these screens presents significant statistical challenges. For example, technical factors like sequencing depth impact not only expression measurement but also perturbation detection, creating a confounding effect. We demonstrate on a recent large-scale single cell CRISPR screen how these challenges cause calibration issues among existing analysis methods. To address these challenges, we propose SCEPTRE: analysis of single cell perturbation screens via conditional resampling. This methodology, designed to avoid calibration issues due to technical confounders and expression model misspecification, infers associations between perturbations and expression by resampling the former according to a working model for perturbation detection probability in each cell. Applying SCEPTRE to the CRISPR screen data yields excellent calibration on negative control perturbations and 217 regulatory relationships not found in the original study, many of which are supported by existing functional data.
\end{abstract}

@@

To address these challenges, we propose SCEPTRE (analysis of Single CEll PerTurbation screens via conditional REsampling). This resampling methodology is based on the conditional randomization test\cite{CetL16}, a recently proposed statistical methodology which, like traditional nonparametric methods, does not rely on correct specification of the outcome (expression) model. Unlike traditional nonparametric methods, SCEPTRE employs a resampling scheme that explicitly accounts for heterogeneity in technical factors among cells. We analyze both synthetic data and that of Gasperini et al.\cite{Gasperini2019} to illustrate the aforementioned analysis challenges and the performance of the proposed methodology. SCEPTRE demonstrates excellent calibration and reveals many novel regulatory relationships supported by a variety of existing sources of functional evidence.

@@

The Gasperini et al.\cite{Gasperini2019} data exemplify several of the analysis challenges in single cell CRISPR screens. A library of CRISPR guide RNAs (gRNAs) targeting 5,779 candidate enhancers, 50 negative controls, and 381 positive controls was transduced via CROP-seq\cite{Datlinger2017,Hill2018} at high multiplicity of infection into a population of 207,324 K562 cells expressing the Cas9-KRAB repressive complex, each cell receiving an average of 28 perturbations. Polyadenylated gRNA barcodes were sequenced alongside the whole transcriptome, and perturbation identities assigned to cells by thresholding the resulting gRNA UMI counts. Initially, $p$-values for each gRNA-gene pair were obtained from a DESeq2\cite{Love2014}-inspired negative binomial regression analysis implemented in Monocle2\cite{Qiu2017}. By examining the distribution of negative binomial $p$-values pairing each gene with each of 50 non-targeting control (NTC) gRNAs, Gasperini et al. found that these $p$-values are inflated (Figure~\ref{fig:analysis-challenges}a). To remedy this issue, Gasperini et al. calibrated the candidate enhancer $p$-values against the distribution of all gene-NTC $p$-values instead of the uniform distribution. The resulting ``empirical" $p$-values were used for determining significance.

First, we illustrate the confounding effect discussed in the introduction. Despite the use of a targeted amplification protocol for gRNA detection\cite{Hill2018}, not all gRNAs are detected. Indeed, the total number of gRNAs detected in a cell increases with the sequencing depth, defined as the total number of mRNA UMIs detected per cell ($\rho = 0.35, p < 10^{-15}$; Figure~\ref{fig:analysis-challenges}b). We observed a similar phenomenon in the Mosaic-seq data\cite{Xie2017} ($\rho = 0.23, p < 10^{-15}$; Figure~\ref{fig:mosaicseq}). Therefore, sequencing depth and other technical factors induce correlations between the measured gRNA presence and gene expression, even in the absence of a regulatory relationship (Figure~\ref{fig:analysis-challenges}c). Parametric regression approaches like those employed by Gasperini et al. are the easiest way to adjust for confounders, but their heavy reliance on correct model specification creates a different set of calibration issues.

We hypothesized that imperfect dispersion estimation may be responsible for the inflation observed by Gasperini et al. Monocle2 fits a raw dispersion for each gene, then fits a mean-dispersion relationship across genes, and finally collapses each raw dispersion estimate onto this fitted line (Figure~\ref{fig:analysis-challenges}d). While the raw dispersion estimates are noisy and some shrinkage is helpful\cite{Hafemeister2019}, the collapsing of the dispersion estimates to the fitted curve is likely to underestimate the dispersions for points significantly above the curve. On the other hand, points near the curve may not have this issue. Indeed, plotting the 50 raw NTC $p$-values for genes near the curve, we found good calibration (e.g., LUC7L2; Figure~\ref{fig:analysis-challenges}e), while genes significantly above the curve showed marked signs of raw $p$-value inflation (e.g., HIST1H1D; Figure~\ref{fig:analysis-challenges}f).

This heterogeneity in inflation among gene-gRNA pairs reveals a weakness of the ``one-size-fits-all" NTC-based calibration approach employed by Gasperini et al.: building a null distribution using all gene-NTC pairs leads to overcorrection for some gene-enhancer pairs (false negatives) and undercorrection for others (false positives). For example, the empirical correction is unnecessary for genes like LUC7L2. However, we pay a price in sensitivity for this empirical correction; it decreases the significance of the $p$-value for a positive control gRNA by three orders of magnitude (Figure~\ref{fig:analysis-challenges}e). On the other hand, the empirical correction is insufficient for high-dispersion genes. We computed the deviation from uniformity of the empirical NTC $p$-values for each gene using the Kolmogorov-Smirnov (KS) test, represented by the color of each point in Figure~\ref{fig:analysis-challenges}b. Circled genes have significantly miscalibrated empirical $p$-values based on a Bonferroni correction at level $\alpha = 0.05$; 21 of these were flagged by Gasperini et al. as outliers but 21 were not. Among the latter group is HIST1H1D, whose empirical $p$-values are still inflated (Figure~\ref{fig:analysis-challenges}f), leading to two potential false discoveries, one of which is deemed ``high confidence" (Figure~\ref{fig:analysis-challenges}g, Figure~\ref{fig:potential-false-positives}).

\addtocounter{page}{-1}
\thispagestyle{empty} 
\begin{figure}[h!]
	\includegraphics[width = \textwidth]{figures/Figure1/Figure1.png}
	\caption{\textbf{CRISPR screen analysis challenges can lead to false positives and false negatives.} \textbf{a}, QQ-plot of Gasperini et al. $p$-values for all gene-gRNA pairs involving either candidate enhancer or negative control gRNAs. Inflation in negative controls makes it harder to interpret associations between genes and candidate enhancers, motivating the empirical $p$-value correction. \textbf{b-c}, Sequencing depth (total mRNA UMIs per cell) impacts gRNA detection (b) and observed expression levels, and therefore acts as a confounder (c). This analysis challenge distinguishes CRISPR screens from traditional differential expression applications. }
	\label{fig:analysis-challenges}
\end{figure}

@@

We attempted to alleviate the miscalibration within the negative binomial regression framework (Figure \ref{fig:calibration-comparison-all}). First, we replaced the collapsed dispersion estimates by the raw estimates (recall Figure~\ref{fig:analysis-challenges}d). This simple modification is based on the intuition that not as much dispersion shrinkage is necessary as in bulk RNA-seq, since in scRNA-seq we have many more samples (one for each cell) to estimate this parameter. This dispersion estimate modification yielded greatly improved calibration over all gene-NTC pairs (blue curve in Figure \ref{fig:calibration-comparison-all}a), but separating by NTC revealed substantial remaining gRNA-dependent miscalibration (Figure~\ref{fig:calibration-comparison-all}b). 
This dependency on gRNA further underscores the problem with ``one-size-fits-all" NTC-based calibration approaches, also implemented by Xie et al.\cite{Xie2019}

We hypothesized that this remaining miscalibration stems from inadequate sequencing depth correction. Indeed, we found we could predict the direction and strength of the miscalibration for each gRNA from an adjusted correlation between the sequencing depth and the gRNA presence (Methods). This yielded $z$-values as shown in the legend of Figure~\ref{fig:calibration-comparison-all}b; large positive (negative) $z$-values signal spurious positive (negative) correlations between gRNA presence and sequencing depth, leading to conservative (liberal) bias. Gasperini et al. corrected for sequencing depth using DESeq2-style size factors, whereas recent work on scRNA-seq analysis\cite{Hafemeister2019} advocates for including sequencing depth as a covariate in the negative binomial regression. We therefore included as covariates both the total number of UMIs observed and the number of genes with at least one UMI in a given cell, in addition to the technical factors Gasperini et al. corrected for (see Methods).

To assess the impact of these two improvements, we applied the negative binomial test with and without these improvements (four total possibilities) to all 50 NTC gRNAs paired with all genes  (Figure~\ref{fig:calibration-comparison-all}a,c). The changes to confounder correction and dispersion estimation both markedly improve calibration, especially when used in conjunction. However, the resulting improved negative binomial method still shows clear signs of miscalibration, producing mostly conservative $p$-values. There are many possible reasons for the remaining miscalibration, most likely due to remaining misspecification of the negative binomial model. While more effort could certainly be invested in further improvements of this parametric model for gene expression, we acknowledge that modeling scRNA-seq expression is a challenging problem that remains open. This highlights the appeal of nonparametric approaches; however, as discussed before, traditional nonparametric approaches are not applicable in the presence of technical factors impacting both gRNA detection and gene expression measurement. To resolve this impasse, we propose a resampling-based approach designed to account for such confounders.

\paragraph{SCEPTRE: Analysis of single cell perturbation screens via conditional resampling.}

@@

\clearpage
\thispagestyle{empty} 
\addtocounter{page}{-1}
\begin{figure}[h!]
	\includegraphics[width = \textwidth]{figures/Figure2/Figure2.png}
	\caption{\textbf{SCEPTRE: Analysis of single cell perturbation screens via conditional resampling.} A schematic and outline of the SCEPTRE methodology for one gene and one gRNA is shown, entailing the application of the conditional randomization test\cite{CetL16} to single cell CRISPR screens. The idea is to assign to every cell the probability of gRNA observation based on its technical factors, and then to build a null distribution for the negative binomial $z$-value by independently resampling the gRNA indicator for each cell according to these probabilities to form ``negative control" data sets. A skew-$t$ distribution is fit to the resulting histogram to obtain precise $p$-values based on a limited number of resamples, against which the original NB $z$-value is compared. The dashed line shows the standard normal distribution, against which the NB $z$-value would normally be compared.}
	\label{fig:method}
\end{figure}
\clearpage

\paragraph{SCEPTRE is well calibrated despite sequencing depth confounding and expression model misspecification.}

First, we demonstrated SCEPTRE's calibration in a small proof-of-concept simulation with 1,000 cells, one gene, one NTC gRNA, and one confounder (Figure~\ref{fig:calibration}a). We considered the one-sided $z$ test statistic based on a negative binomial regression with dispersion 1. We then generated the data from this model as well as from three others: one with dispersion 0.2, one with dispersion 5, and one with dispersion 1 but with 25\% zero-inflation. We compared three ways of building a null distribution for the test statistic based on the possibly misspecified negative binomial model: the standard parametric approach, the permutation approach (based on permuting gRNA assignments), and conditional resampling. The parametric approach works as expected when the negative binomial model is correctly specified, but breaks down in all three cases of model misspecification. The permutation approach is systematically conservative because of inadequate confounder correction. Finally, SCEPTRE is well-calibrated regardless of expression model misspecification.

Next, we applied SCEPTRE to test the association between all NTC gRNAs and all genes in the Gasperini et al. data (Figure~\ref{fig:calibration}b-c). For comparison, we also applied the original negative binomial method, the improved negative binomial method, and a permutation-based calibration of the latter. The simplest calibration assessment is to compare the resulting 538,560 $p$-values to the uniform distribution (Figure~\ref{fig:calibration}b). SCEPTRE shows excellent calibration, substantially improving on the parametric approach based on the same test statistic. Breaking the pairs down by NTC (Figure~\ref{fig:calibration}c), SCEPTRE again shows nearly perfect calibration, as evidenced by its colored confidence region coinciding with the gray shaded region representing the 95\% confidence band for the uniform distribution.

To assess the sensitivity of SCEPTRE, we applied it to the 381 positive control gRNA / gene pairs assayed in Gasperini et al (Figure~\ref{fig:calibration}d), where each gRNA targeted the transcription start site of the corresponding gene. Comparing to the original empirical $p$-values, SCEPTRE $p$-values for these positive controls are generally more significant, suggesting greater sensitivity. We note that the empirical correction employed by Gasperini et al. limits the accuracy of $p$-values to about $10^{-6}$, explaining at least part of the difference.

\clearpage
\thispagestyle{empty} 
\addtocounter{page}{-1}
\begin{figure}[h!]
	\includegraphics[width = \textwidth]{figures/Figure3/Figure3.png}
	\caption{\textbf{SCEPTRE improves on the sensitivity and specificity of the parametric approach by correcting for expression model misspecification.} \textbf{a,} Numerical simulation comparing three ways of calibrating a negative binomial test statistic (parametric, permutation, conditional resampling), when the test statistic is based on the correct model or contains expression model misspecifications. Only the conditional resampling approach maintains calibration despite model misspecification. \textbf{b,c,} Application of SCEPTRE to Gasperini et al negative control gRNAs, comparing all gene-NTC pairs to the uniform distribution (b) or breaking down by NTC (c); details as in Figure~\ref{fig:calibration-comparison-all}a,c. The matching of SCEPTRE's colored region with the gray one in panel c, together with the overall near-uniformity in panel b, demonstrates its excellent calibration on real negative control data. \textbf{d,} SCEPTRE $p$-values for TSS-targeting controls generally more significant than the original empirical $p$-values; dashed line is the $45^\circ$ line. Note: $p$-values truncated for visualization, and empirical $p$-values in panel d at least 1/538,560 $\approx$ 2$\times 10^{-6}$ by construction.}
	\label{fig:calibration}
\end{figure}
\clearpage

@@

We applied SCEPTRE to the 84,595 gene-enhancer pairs considered in Gasperini et al., encompassing 10,560 genes and 5,779 candidate enhancers (Figure~\ref{fig:application}a). We applied the Benjamini-Hochberg correction at level 0.1 (the same level used in the original analysis) to the $p$-values obtained for all of these candidate pairs, obtaining a total of 585 gene-enhancer pairs. By comparison, Gasperini et al. found 470 high-confidence pairs. Comparing the SCEPTRE $p$-values against the original empirical $p$-values (Figure~\ref{fig:application}b), we see that the two often diverge substantially. Our analysis found 217 gene-enhancer pairs that the original analysis did not, while 102 were found only by the original analysis. Many of the discoveries found only in the original analysis show signs of the $p$-value inflation observed in Figure~\ref{fig:analysis-challenges} (Figure~\ref{fig:potential-false-positives}).

Among the 217 new gene-enhancer pairs discovered, several are supported by evidence from orthogonal functional assays. In particular, we highlight five of these pairs (Figure~\ref{fig:application}c) involving genes not paired to any enhancers in the original analysis, which are supported by GTEx\cite{GTExPilot} eQTL $p$-values in whole blood or enhancer RNA correlation $p$-values across tissues from the FANTOM project\cite{Andersson2014}. These pairs are listed in the GeneHancer database\cite{Fishilevich2017}, which aggregates eQTL, eRNA, and other sources of evidence of gene-enhancer interactions. The SCEPTRE $p$-values for these promising pairs are generally 1-2 orders of magnitude more significant than the original empirical $p$-values. Also among the 217 new gene-enhancer pairs are 9 pairs involving 7 genes that were deemed outliers in the original analysis (blue triangles in Figure~\ref{fig:application}b), underscoring SCEPTRE's ability to handle genes with arbitrary distributions that may not fit into traditional parametric models.

>>>>

\usepackage{graphicx}	
\usepackage{amssymb, amsmath, bbm}
\usepackage{geometry}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\newcommand\independent{\perp \!\!\! \perp}
% \usepackage{chngcntr}

@@

\begin{abstract}
Single cell CRISPR screens are an emerging biotechnology promising unprecedented insights into gene regulation. However, the analysis of these screens presents significant statistical challenges. For example, technical factors like sequencing depth impact not only expression measurement but also perturbation detection, creating a confounding effect. We demonstrate on two recent large-scale single cell CRISPR screens how these challenges cause calibration issues among existing analysis methods. To address these challenges, we propose SCEPTRE: analysis of single cell perturbation screens via conditional resampling. This methodology, designed to avoid calibration issues due to technical confounders and expression model misspecification, infers associations between perturbations and expression by resampling the former according to a working model for perturbation detection probability in each cell. Applying SCEPTRE to the CRISPR screen data yields excellent calibration on negative control perturbations and 200 regulatory relationships not found in the original study, many of which are supported by existing functional data.
\end{abstract}

@@

To address these challenges, we propose SCEPTRE (analysis of Single CEll PerTurbation screens via conditional REsampling; pronounced ``scepter''). This resampling methodology is based on the conditional randomization test\cite{CetL16}, a recently proposed statistical methodology which, like traditional nonparametric methods, does not rely on correct specification of the outcome (expression) model. Unlike traditional nonparametric methods, SCEPTRE employs a resampling scheme that explicitly accounts for heterogeneity in technical factors among cells. We analyze three datasets -- that of Gasperini et al.\cite{Gasperini2019}, that of Xie et al.\cite{Xie2019}, and a synthetic dataset -- to illustrate the aforementioned analysis challenges and the performance of the proposed methodology. SCEPTRE demonstrates excellent calibration and reveals many novel regulatory relationships supported by a variety of existing sources of functional evidence.

@@

The Gasperini et al.\cite{Gasperini2019} and Xie et al.\cite{Xie2019} data exemplify several of the analysis challenges in single cell CRISPR screens. Gasperini et al.\ transduced via CROP-seq\cite{Datlinger2017,Hill2018}  at a high multiplicity of infection a library of CRISPR guide RNAs (gRNAs) into a population of 207,324 K562 cells expressing the Cas9-KRAB repressive complex, each cell receiving an average of 28 perturbations. The gRNA library targeted 5,779 candidate enhancers, 50 negative controls, and 381 positive controls. Xie et al.\ used Mosaic-seq\cite{Xie2019,Xie2017} to perturb at a high multiplicity of infection 518 putative enhancers in a population of 106,670 Cas9-KRAB-expressing K562 cells. Each putative enhancer was perturbed in an average of 1,276 cells. Both Gasperini et al.\ and Xie et al.\ sequenced polyadenylated gRNA barcodes alongside the whole transcriptome and assigned perturbation identities to cells by thresholding the resulting gRNA UMI counts.

Gasperini et al.\ and Xie et al.\ computed $p$-values for gRNA-gene pairs
to test for associations between candidate enhancers and genes. Gasperini et al.\ obtained $p$-values from a DESeq2\cite{Love2014}-inspired negative binomial regression analysis implemented in Monocle2.\cite{Qiu2017} Xie et al., in contrast, obtained $p$-values from Virtual FACS, a nonparametric rank-based method proposed by the authors.\cite{Xie2017} Gasperini et al.\ and Xie et al.\  observed miscalibration in their respective sets of $p$-values. Gasperini et al.\ examined the distribution of negative binomial $p$-values obtained from pairing each protein-coding gene with each of 50 non-targeting control gRNAs. These ``null'' $p$-values exhibited inflation, deviating substantially from the expected uniform distribution (Figure \ref{fig:analysis-challenges}a, top panel). Xie et al.\ examined the distribution of Virtual FACS $p$-values obtained from pairing gRNAs that targeted an enhancer of gene \textit{ARL15}, ARL15-enh, with all protein-coding genes and long noncoding RNAs (lncRNAs) genome-wide. As ARL15-enh is expected to modulate the expression of ARL15 and neighboring lncRNAs only, these $p$-values can be used to assess model calibration. Xie et al.\
 found these $p$-values to be miscalibrated, with those corresponding to protein-coding genes exhibiting conservative bias (Figure \ref{fig:analysis-challenges}a, bottom panel) and those corresponding to lncRNAs exhibiting liberal bias (not depicted).

Confounding likely is a cause of the miscalibration noted by Gasperini et al.\ and Xie et al. Despite the use of targeted amplification protocols for gRNA detection\cite{Hill2018}, not all transduced gRNAs are detected in single cell CRISPR screen experiments. Indeed, the total number of gRNAs detected in a cell increases with the sequencing depth, defined as the total number of mRNA UMIs detected per cell ($\rho = 0.35, p < 10^{-15}$ in Gasperini et al.\ data; $\rho = 0.25, p < 10^{-15}$ in Xie et al.\ data; Figures \ref{fig:analysis-challenges}b-c). Therefore, sequencing depth and other technical factors induce correlations between the measured gRNA presence and gene expression, even in the absence of a regulatory relationship (Figure \ref{fig:analysis-challenges}d). This confounding effect, if not explicitly accounted for, results in test miscalibration. Confounding is especially problematic for nonparametric methods like Virtual FACS, which implicitly (and incorrectly) assume that there are no technical factors that impact both gRNA detection and gene expression measurement.

Parametric regression approaches, like those employed by Gasperini et al., are the easiest way to adjust for confounders. However, parametric methods rely heavily on correct model specification, a challenge in single cell analysis given the heterogeneity, complexity, and occasional multimodality of the data. We hypothesized that inaccurate estimation of the negative binomial dispersion parameter was (in part) responsible for the $p$-value inflation observed by Gasperini et al. Monocle2 fits a raw dispersion for each gene, then fits a parametric mean-dispersion relationship across genes, and finally collapses each raw dispersion estimate onto this fitted line (Figure \ref{fig:analysis-challenges}e). We computed the deviation from uniformity of the negative control $p$-values for each gene using the Kolmogorov-Smirnov (KS) test, represented by the color of each point in Figure \ref{fig:analysis-challenges}e. Circled genes have significantly miscalibrated $p$-values based on a Bonferroni correction at level $\alpha = 0.05$. Genes significantly above the curve showed marked signs of $p$-value inflation.

Gasperini et al.\ and Xie et al.\ adjusted their analyses to remedy the observed calibration problems. Gasperini et al.\ calibrated the candidate enhancer $p$-values against the distribution of all gene-negative control $p$-values instead of the uniform distribution. The resulting ``empirical $p$-values'' were used for determining significance. Along similar lines, Xie et al.\ calibrated the Virtual FACS $p$-values against a simulated set of gene-specific ``null'' $p$-values. These analysis adjustments, unfortunately, did not satisfactorily improve calibration. The adjustment of Gasperini et al., for example, lead to overcorrection for some gene-enhancer pairs (false negatives) and undercorrection for others (false positives) (Figure 1s). Methods that avoid reliance on ad hoc calibration fixes might produce more consistently calibrated $p$-values across entire CRISPR screen datasets.

\begin{figure}[h!]
\includegraphics[width = \textwidth]{figures/Figure1/Figure1_cropped.pdf}
\caption{\textbf{CRISPR screen analysis challenges can lead to false positives and false negatives}. \textbf{a}, (Top) QQ-plot of Gasperini et al.\ p-values for all pairs of genes and negative control gRNAs. (Bottom) QQ-plot of Xie et al.\ p-values for all pairs of genes and ARL15-enh-targeting gRNAs (\textit{ARL15} gene omitted). These sets of $p$-values deviate from the expected uniform distribution, indicating test miscalibration. \textbf{b-d}, Sequencing depth (total mRNA UMIs per cell) impacts gRNA detection and observed gene expression levels in both Gasperini et al.\ (\textbf{b}) and Xie et al.\ (\textbf{c}) data. Thus, sequencing depth acts as a confounder (\textbf{d}). This analysis challenge distinguishes CRISPR screens from traditional differential expression applications. \textbf{e}, Monocle2 estimates the dispersion of each gene by projecting each gene's raw dispersion estimate onto the fitted raw dispersion-mean expression curve. This estimation procedure leads to miscalibration for high-dispersion genes.  % \textbf{b-c}, Sequencing depth (total mRNA UMIs per cell) impacts gRNA detection (b) and observed expression levels, and therefore acts as a confounder (c). This analysis challenge distinguishes CRISPR screens from traditional differential expression applications. }
}
 \label{fig:analysis-challenges}
\end{figure}
\clearpage

@@

We attempted to alleviate the miscalibration within the negative binomial regression framework. To this end, we followed the recommendations of Hafmeister and Satija, who recently proposed a strategy for parametric modeling of single cell RNA-seq data that, to the best of our knowledge, is state-of-the-art.\cite{Hafemeister2019}  First, we abandoned the DESeq2-style size factors of Monocle2 and instead corrected for sequencing depth by including sequencing depth as a covariate in the negative binomial regression model. Second, we adopted a more flexible dispersion estimation procedure. Following Hafmeister and Satija, we (i) computed raw dispersion estimates for each gene, (ii) regressed the raw dispersion estimates onto the mean gene expressions via kernel regression, and (iii) projected the raw dispersion estimates onto the fitted nonparametric regression curve.

We reanalyzed the Gasperini et al.\ data using the improved negative binomial regression approach. In addition to sequencing depth, we included as covariates in the regression model the total number of expressed genes per cell, as well as the technical factors that Gasperini et al.\ accounted for in their analysis (namely, total number of gRNAs detected per cell, percentage of transcripts mapped to mitochondrial genes, and sequencing batch). We also applied the improved negative binomial approach to the Xie et al.\ ARL15-enh data. In the latter analysis we included sequencing depth, number of detected gRNA UMIs per cell, and sequencing batch as covariates. We found that the improved negative binomial model was better calibrated than the Monocle2 negative binomial model on the Gasperini data (Figure 3c; improved negative binomial denoted ``Hafemeister NB''). However, on both the Gasperini et al.\ data and Xie et al.\ data, the improved negative binomial model showed clear signs of p-value inflation (Figure 3c-d).

There are many possible explanations for the remaining miscalibration, mostly due to misspecification of the negative binomial model. While more effort could be invested into further improvements of a parametric model for gene expression, modeling single cell data remains a challenging problem outside the scope of the current work. This difficulty highlights the appeal of nonparametric methods for singe cell analysis.\cite{Zhu2019} However, as discussed in the previous section, traditional nonparametric approaches are not applicable in the presence of technical factors that impact both gRNA detection and gene expression measurement. To resolve this impasse, we propose a resampling-based approach designed to account for such confounders.

% The difficulty of applying parametric models to single cell data highlights the appeal of nonparametric approaches for this purpose.
\clearpage
\paragraph{SCEPTRE: Analysis of single cell perturbation screens via conditional resampling.}

@@

\clearpage
\thispagestyle{empty}
\begin{figure}[h!]
	\includegraphics[width = \textwidth]{figures/Figure2/Figure2.png}
	\caption{\textbf{SCEPTRE: Analysis of single cell perturbation screens via conditional resampling.} A schematic and outline of the SCEPTRE methodology for one gene and one gRNA is shown, entailing the application of the conditional randomization test\cite{CetL16} to single cell CRISPR screens. The idea is to assign to every cell the probability of gRNA observation based on its technical factors, and then to build a null distribution for the negative binomial $z$-value by independently resampling the gRNA indicator for each cell according to these probabilities to form ``negative control" data sets. A skew-$t$ distribution is fit to the resulting histogram to obtain precise $p$-values based on a limited number of resamples, against which the original NB $z$-value is compared. The dashed line shows the standard normal distribution, against which the NB $z$-value would normally be compared.}
	\label{fig:method}
\end{figure}
\clearpage

\paragraph{SCEPTRE produces correct results when the ground truth is known.}

We assessed the calibration, sensitivity, and specificity of SCEPTRE in several settings in which the ground truth was known. First, we demonstrated the calibration of SCEPTRE in a small, proof-of-concept simulation study (Figure \ref{fig:ground_truth}a). We considered a class of negative binomial regression models with dispersion $\alpha$ and two technical covariates (sequencing depth and batch). We simulated expression data for a single gene in 1000 cells using four models selected from this class: the first with $\alpha = 1$, the second with $\alpha = 0.25$, the third with $\alpha = 5$, and the last with $\alpha = 1,$ but with 25$\%$ zero-inflation. We also simulated negative control gRNA data using a logistic regression model with the same covariates as in the gene expression model. We assessed the calibration of three methods across the four simulated datasets: SCEPTRE, improved negative binomial regression, and scMAGeCK-LR,\cite{Yang2020} a recently-proposed, permutation-based nonparametric method. To assess the impact of model misspecification on SCEPTRE and the improved negative binomial method (on which SCEPTRE relies), we fixed the dispersion of the negative binomial method to 1 across all four simulated datasets. We found that the negative binomial method works as expected when the model is correctly specified but breaks down in all three cases of model misspecification. scMAGeCK-LR exhibits poor calibration across all simulated datasets, likely because, as a traditional nonparametric method, it fails to adequately account for confounders. Finally, SCEPTRE is well-calibrated regardless of expression model misspecification and confounder presence.

Next, to assess the calibration of SCEPTRE on real data, we applied SCEPTRE to test the association between all negative control gRNAs and all genes in the Gasperini et al.\ data (Figure \ref{fig:ground_truth}b) and all ARL15-enh-targeting gRNAs and all genes in the Xie et al.\ data (Figure \ref{fig:ground_truth}c). We compared SCEPTRE to the improved negative binomial method, as well as to the original analysis methods (i.e., Monocle regression on the Gasperini et al.\ data and Virtual FACS on the Xie et al.\ data). We did not apply scMAGeCK-LR to these data given its poor calibration on the simulated data. SCEPTRE shows excellent calibration on the Gasperini et al.\ negative control data; Monocle regression and improved negative binomial regression, in contrast, demonstrate signs of severe $p$-value inflation. Similarly, on the Xie et al.\ ARL15-enh data, SCEPTRE exhibits nearly perfect calibration, whereas Virtual FACS and the improved negative binomial method show conservative and liberal bias, respectively.

Last, we assessed the sensitivity of SCEPTRE using the Gasperini et al.\ positive control data and confirmatory bulk RNA-seq data produced by Xie et al. We applied SCEPTRE to the 381 positive control gRNA-gene pairs assayed in Gasperini et al.\ (Figure~\ref{fig:ground_truth}d), where each gRNA targeted the transcription start site of the corresponding gene. The SCEPTRE $p$-values for these positive controls are highly significant, and in particular, more significant than the original empirical $p$-values, indicating greater sensitivity. We note that the empirical correction employed by Gasperini et al.\ limits the accuracy of $p$-values to about $10^{-6}$, explaining at least part of the difference. Finally, we compared the SCEPTRE results for ARL15-enh to the results of an arrayed CRISPR screen of ARL15-enh in which populations of perturbed cells were compared to populations of unperturbed cells via bulk RNA-seq.\cite{Xie2019} Both SCEPTRE and the bulk RNA-seq differential expression analysis rejected gene \textit{ARL15} (and only gene \textit{ARL15}) at an FDR of 0.1 after a Benjamini-Hochberg correction, increasing our confidence in sensitivity (and specificity) of SCEPTRE (Figure \ref{fig:ground_truth}e).

\newpage
\begin{figure}[h!]
	\includegraphics[width = \textwidth]{figures/Figure3/Figure3_cropped.pdf}
	\caption{\textbf{SCEPTRE demonstrates excellent calibration, specificity, and sensitivity when the ground truth is known.}}
	\label{fig:ground_truth}
 \end{figure}
\clearpage

% \clearpage
% \thispagestyle{empty} 
% \addtocounter{page}{-1}
%\begin{figure}[h!]
%	\includegraphics[width = \textwidth]{figures/Figure3/Figure3.png}
%	\caption{\textbf{SCEPTRE improves on the sensitivity and specificity of the parametric approach by correcting for expression model misspecification.} \textbf{a,} Numerical simulation comparing three ways of calibrating a negative binomial test statistic (parametric, permutation, conditional resampling), when the test statistic is based on the correct model or contains expression model misspecifications. Only the conditional resampling approach maintains calibration despite model misspecification. \textbf{b,c,} Application of SCEPTRE to Gasperini et al negative control gRNAs, comparing all gene-NTC pairs to the uniform distribution (b) or breaking down by NTC (c); details as in Figure~\ref{fig:calibration-comparison-all}a,c. The matching of SCEPTRE's colored region with the gray one in panel c, together with the overall near-uniformity in panel b, demonstrates its excellent calibration on real negative control data. \textbf{d,} SCEPTRE $p$-values for TSS-targeting controls generally more significant than the original empirical $p$-values; dashed line is the $45^\circ$ line. Note: $p$-values truncated for visualization, and empirical $p$-values in panel d at least 1/538,560 $\approx$ 2$\times 10^{-6}$ by construction.}
%	\label{fig:calibration}
% \end{figure}
\clearpage

@@

We applied SCEPTRE to the 84,595 gene-enhancer pairs considered in Gasperini et al., encompassing 10,560 genes and 5,779 candidate enhancers. We applied the Benjamini-Hochberg correction at level 0.1 (the same level used in the original analysis) to the $p$-values obtained for all of these candidate pairs, obtaining a total of 563 gene-enhancer pairs. By comparison, Gasperini et al. found 470 high-confidence pairs. Comparing the SCEPTRE $p$-values against the original empirical $p$-values (Figure~\ref{fig:application}a), we see that the two often diverge substantially. Our analysis found 200 gene-enhancer pairs that the original analysis did not, while 107 were found only by the original analysis. Many of the discoveries found only in the original analysis show signs of the $p$-value inflation observed in Figure~\ref{fig:analysis-challenges}.

Among the 200 new gene-enhancer pairs discovered, several are supported by evidence from orthogonal functional assays. In particular, we highlight five of these pairs (Figure~\ref{fig:application}b) involving genes not paired to any enhancers in the original analysis, which are supported by GTEx\cite{GTExPilot} eQTL $p$-values in whole blood or enhancer RNA correlation $p$-values across tissues from the FANTOM project\cite{Andersson2014}. These pairs are listed in the GeneHancer database\cite{Fishilevich2017}, which aggregates eQTL, eRNA, and other sources of evidence of gene-enhancer interactions. The SCEPTRE $p$-values for these promising pairs are generally 1-2 orders of magnitude more significant than the original empirical $p$-values. Also among the 200 new gene-enhancer pairs are 6 pairs involving 5 genes that were deemed outliers in the original analysis (blue triangles in Figure~\ref{fig:application}a), underscoring SCEPTRE's ability to handle genes with arbitrary distributions that may not fit into traditional parametric models.

We also found that the total set of gene-enhancer pairs discovered was better enriched for regulatory biological signals, including HI-C and ChIP-seq. Gene-enhancer pairs discovered by SCEPTRE tended to be physically closer to each other on average that those discovered in the original analysis (Figure~\ref{fig:application}c). Furthermore, 74.4\% of SCEPTRE's 563 gene-enhancer pairs fell in the same topologically associating domain (TAD), compared to 71.1\% of the original 470 pairs. We also repeated Gasperini et al.'s contact frequency enrichment analysis for those pairs falling in the same TAD (Figure~\ref{fig:application}d). We found similar levels of contact frequency enrichment, despite the fact that SCEPTRE discovered 85 more gene-enhancer pairs in the same TADs. Finally, we repeated the ChIP-seq enrichment analysis of Gasperini et al. for seven cell-type relevant transcription factors and H3K27ac. We quantified enrichment of ChIP-seq signal in paired enhancers as the odds ratio that a candidate enhancer is paired to a gene, comparing those falling in the top quintile of candidate enhancers by ChIP-seq signal and those not overlapping a ChIP-seq peak at all. We find improved enrichment for each of the eight ChIP-seq targets (Figures~\ref{fig:application}e and \ref{fig:chipseq-enrichment-details}).
